Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-02-15 paper_comparison/paper_comparison/paper_comparison.Rmd
update the text according to the updated effect size

We make an assumption that all the realizations of $\boldsymbol{X}$ are generated from a multivariate normal distribution $N(\boldsymbol{0}, 0.3^2\boldsymbol{I})$ and then rescaled by the min-max normalization. The estimate of $E$ is obtained by taking the average of the $E_{sample}$ calculated for samples drawn from the given model $M$ repeatedly.

@@

To get an accurate estimate of the power of conventional tests, we simulate 200000 samples evenly from the models used in the human subject experiments, and record the $p$-values produced by every sample.

@@

For non-linearity patterns, the power curve of RESET test climbs aggressively from
`r round(unname(predict(reset_poly_glmmod, newdata = data.frame(effect_size = exp(0), offset0 = log(0.05/0.95)), type = "response")) * 100)`% to `r round(unname(predict(reset_poly_glmmod, newdata = data.frame(effect_size = exp(2.5), offset0 = log(0.05/0.95)), type = "response")) * 100)`% as $log_e(E)$ increases from 0 to 2.5, while power of other tests respond inactively to the change of effect, showing that RESET test is way more sensitive to the type of model defects that being considered. Meanwhile, no noticeable residual depatures can be spotted from the example residual plots.

In terms of heteroskedasticity patterns, the power of Breusch-Pagan test is also almost always greater than the power of visual test. For $log_e(E)$ at around 2.5, where the power curve of the visual test remains at a low level, the Breusch-Pagan test still have a decent amount of chance of rejecting $H_0$. Similarly, the visual feature is nearly unobservable from the example residual plots.

@@

```{r polypower, fig.height = 3, fig.width = 5, fig.cap = "Comparison of power between different tests for non-linear patterns (uniform fitted values only). Main plot shows the power curves estimated using logistic regression, with dots indicating human evaluations of lineups. Surrounding lines of the visual test show the estimated power of 500 bootstrap samples. Small row of plots shows typical residual plots corresponding to specific effect sizes, marked by dashed lines in main plot. Where would you draw the line of too much non-linearity in the residuals? For the RESET test this is around log effect size 2.5, but for the visual test it is around 3.5."}

@@

```{r heterpower, fig.height = 3, fig.width = 5, fig.cap = "Comparison of power between different tests for heteroskedasticity patterns (uniform fitted values only). Main plot shows the power curves, with dots indicating human evaluations of lineups. Surrounding lines of the visual test show the estimated power of 1000 bootstrap samples. Small row of plots shows typical residual plots corresponding to specific effect sizes, marked by dashed lines in main plot. Where would you draw the line of too much heteroskedasticity in the residuals? For the BP test this is around log effect size 2.5, but for the visual test it is around 3.5."}

@@

```{r heter-example, fig.height = 7, fig.width = 7, fig.cap = 'The single heteroskedasticity lineup that is rejected by the visual test but not by the BP test. The data plot at panel 17 contains a "Butterfly" shape. It has effect size $ = 4.02$, somewhat surprising that it is not detected by the BP test.'}
# plot heter-331
VI_MODEL$plot_lineup(vi_lineup$heter_331$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(),
                     remove_axis = TRUE)
```

@@

A primary factor contributes to the non-linearity model is the shape of the non-linearity pattern. According to Figure \ref{fig:poly-power-uniform-j}, conventional tests have identical power in testing the "U", the "S" and the "M" shape. But it has significant lower power in testing the "Triple-U" shape. To understand why this is, one needs to return to the way the RESET test is applied. It requires a parameter indicating degree of fitted values to test for, and the recommendation is to generically use four (@ramsey_tests_1969). However, the "Triple-U" shape constructed from the Hermite polynomials use power up to 18. If the RESET test had been applied using a higher power no less than six, the power curve of "Triple-U" shape will have little difference to other shapes. The recommendation of the polynomial power for the RESET should be revised, perhaps. This illustrates the sensitivity of conventional testing to the parameters, and it also points to a limitation that one needs to know the data generating process in order to set the parameters for the test.

For visual tests, based on the orders of the Hermite polynomials, we expect the "U" shape will be the easiest one to be detected by subjects followed by the "S", "M" and "Triple-U" shape. From Figure \ref{fig:poly-power-uniform-j}, it can be observed that the power curves are mostly aligned with our expectation, except for the "M" shape, which is as easy to be detected as the "U" shape. This implies the readability of the shape do not strictly follow the degree of the polynomials.

@@

(change the butterfly color)

We have also investigated the impact of different heteroskedasticity shapes on power of conventional tests and visual tests. In theory, the "Left-triangle" and the "Right-triangle" shapes are functionally identical from the point of view of a Breusch-Pagan test. As shown in Figure \ref{fig:heter-power-uniform-a}, this is indeed the case where little difference between the power curves can be perceived. Similarly, visual tests should have the smae power in detecting these two shapes if they are equally likely to be identified. However, it can be observed from Figure \ref{fig:heter-power-uniform-a} that the power curve for the "Left-triangle" shape is constantly higher than the one for the "Right-triangle" shape, indicating a potential favour of orientation by human, which is worth to be explored in future studies. Besides, the chance of detecting the "Butterfly" shape by visual tests are slightly higher than other two shapes.

@@

# Manually fit the model
dat_heter_a %>%
  nest(dat = c(effect_size, offset0, reject)) %>%
  mutate(mod = map(dat, 
                   ~glm(reject ~ effect_size - 1, 
                        family = binomial(), 
                        data = .x,
                        offset = offset0))) %>%
  mutate(power = map(mod, function(mod) {
    result <- data.frame(effect_size = seq(min_heter_es, max_heter_es, 0.1),
                         offset0 = log(0.05/0.95))
    result$power <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-dat, -mod) %>%
  unnest(power) %>%
  mutate(log_effect_size = log(effect_size)) %>%
  ggplot() +
  geom_line(aes(log_effect_size, power, col = a), size = 1) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  scale_color_manual(values = rcartocolor::carto_pal(3, "Earth")) +
  theme_light() +
  theme(legend.position = "bottom") +
  facet_wrap(~conv_or_not) +
  labs(col = "Heteroskedasticity shape")
```

@@

For non-linearity patterns, we do not observe significant difference in power of visual test between distributions, except for the discrete uniform distribution. It could be due to the fact that humans have difficulties in recognising visual patterns from discreteness, which makes the shape disconnected and incomplete (ref?). In terms of conventional tests, the discrete uniform distribution is instead the one with the greatest power, followed by lognormal, uniform and the normal distribution.

For heteroskedasticity patterns, the power of visual test on lineups with uniform distribution has the greatest power. This is as expected since other three distributions could reduce the chance of revealing the underlying visual pattern because of uneven data points. Although the power of visual test under the discrete uniform distribution is the lowest in the case of non-linearity patterns, it has a relatively great power this time. Considering for heteroskedasticity, the visual patterns are usually detected by connecting the maximum and minimum residuals separately at different fitted values. It will not be greatly affected by the discrete uniform distribution compared to the uniform distribution. Visual tests have low power on lineups with the normal distribution and the lognormal distribution. This suggests the use of transformation of fitted values in reading residual plots. In the case of conventional tests, the discrete uniform distribution consistently perform well, followed by uniform, normal and the lognormal distribution.

@@

```{r different-x-dist-poly-power, fig.cap = "Comparison of power on lineups with different fitted value distributions, for conventional tests and visual tests. For visual tests, all the power curves are estimated using five evaluations on each lineup. For lineups with uniform fitted value distribution, the five evaluations are randomly sampled from the total eleven evaluations. The sampling process has been repeated for 500 times and the corresponding power curves are drawn with transparent grey lines. For both types of tests power does change. The discrete fitted value distribution substantially lowers the power of visual tests for non-linearity patterns in residual plots. For heteroskedasticity patterns, normal and lognormal distributions produce lower power for both types of tests."}

@@

Now it is important to note that residual plots need to be delivered as a line up, where it is embedded in a field of null plots. This enables a careful calibration for reading structure in residual plots. XXX More here

@@

As expected, if the fitted value distribution is not uniform, there is a loss of power in the visual test. Structure is hardest to detect if fitted values ar discrete. Also, a surprising find was that the direction of heteroskedasticity appears to affect the ability to visually detect it, with wedge to the left being less detectable.

>>>>

We make an assumption that all the realizations of $\boldsymbol{X}$ are generated from a multivariate normal distribution $N(\boldsymbol{0}, 0.3^2\boldsymbol{I})$ and then rescaled by the min-max normalization. The estimate of $E$ is obtained by taking the average of the $E_{sample}$ calculated for samples drawn from the given model $M$ repeatedly. We use the normal distribution as a general choice to approximate the conditional effect size.

@@

@@

For non-linearity patterns, the power curve of RESET test climbs aggressively from
`r round(unname(predict(reset_poly_glmmod, newdata = data.frame(effect_size = exp(0), offset0 = log(0.05/0.95)), type = "response")) * 100)`% to `r round(unname(predict(reset_poly_glmmod, newdata = data.frame(effect_size = exp(2.5), offset0 = log(0.05/0.95)), type = "response")) * 100)`% as $log_e(E)$ increases from 0 to 2.5, while power of other tests respond inactively to the change of effect, showing that RESET test is way more sensitive to the type of model defects that being considered. Meanwhile, no noticeable residual departures can be spotted from the example residual plots.

In terms of heteroskedasticity patterns, the power of Breusch-Pagan test is almost always greater than the power of visual test. For $log_e(E)$ at around 2.5, where the power curve of the visual test remains at a low level, the Breusch-Pagan test has around 50% chance of rejecting $H_0$. Similarly, the visual feature is nearly unobservable from the example residual plots.

@@

```{r polypower, fig.height = 3, fig.width = 5, fig.cap = "Comparison of power between different tests for non-linear patterns (uniform fitted values only). Main plot shows the power curves estimated using logistic regression, with dots indicating human evaluations of lineups. Surrounding lines of the visual test show the estimated power of 500 bootstrap samples. Small row of plots shows typical residual plots corresponding to specific effect sizes, marked by dashed lines in main plot. Where would you draw the line of too much non-linearity in the residuals? For the RESET test this is around log effect size 2.5, but for the visual test it is around 3.5.", dev = 'png', dpi = 450}

@@

```{r heterpower, fig.height = 3, fig.width = 5, fig.cap = "Comparison of power between different tests for heteroskedasticity patterns (uniform fitted values only). Main plot shows the power curves, with dots indicating human evaluations of lineups. Surrounding lines of the visual test show the estimated power of 1000 bootstrap samples. Small row of plots shows typical residual plots corresponding to specific effect sizes, marked by dashed lines in main plot. Where would you draw the line of too much heteroskedasticity in the residuals? For the BP test this is around log effect size 2.5, but for the visual test it is around 3.5.", dev = 'png', dpi = 450}

@@

```{r heter-example, fig.height = 7, fig.width = 7, fig.cap = 'The single heteroskedasticity lineup that is rejected by the visual test but not by the BP test. The data plot at panel 17 contains a "Butterfly" shape. It has effect size $ = 3.76$, somewhat surprising that it is not detected by the BP test.'}
# plot heter-331
VI_MODEL$plot_lineup(vi_lineup$heter_331$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(),
                     remove_axis = TRUE)
```

@@

A primary factor contributes to the non-linearity model is the shape of the non-linearity pattern. According to Figure \ref{fig:poly-power-uniform-j}, conventional tests have higher power in testing shapes constructed from lower orders of Hermite polynomials. But the power in testing the "Triple-U" shape drops significantly. To understand why this is, one needs to return to the way the RESET test is applied. It requires a parameter indicating degree of fitted values to test for, and the recommendation is to generically use four (@ramsey_tests_1969). However, the "Triple-U" shape constructed from the Hermite polynomials use power up to 18. If the RESET test had been applied using a higher power no less than six, the power curve of "Triple-U" shape will only slight lower than the power curve of "M" shape. The recommendation of the polynomial power for the RESET should be revised, perhaps. This illustrates the sensitivity of conventional testing to the parameters, and it also points to a limitation that one needs to know the data generating process in order to set the parameters for the test.

For visual tests, based on the orders of the Hermite polynomials, we expect the "U" shape will be the easiest one to be detected by subjects followed by the "S", "M" and "Triple-U" shape. From Figure \ref{fig:poly-power-uniform-j}, it can be observed that the power curves are mostly aligned with our expectation, except for the "M" shape, which is as easy to be detected as the "S" shape. This implies, unlike the conventional test, the readability of the shape do not strictly follow the degree of the polynomials.

@@

We have also investigated the impact of different heteroskedasticity shapes on power of conventional tests and visual tests. In theory, the "Left-triangle" and the "Right-triangle" shapes are functionally identical from the point of view of a Breusch-Pagan test. As shown in Figure \ref{fig:heter-power-uniform-a}, this is indeed the case where little difference between the power curves can be perceived. Similarly, visual tests should have the same power in detecting these two shapes if they are equally likely to be identified. However, it can be observed from Figure \ref{fig:heter-power-uniform-a} that the power curve for the "Left-triangle" shape is higher than the one for the "Right-triangle" shape, indicating a potential favour of orientation by human, which is worth to be explored in future studies. Besides, the chance of detecting the "Butterfly" shape by both conventional and visual tests are much higher than other two shapes.

@@

# Manually fit the model
dat_heter_a %>%
  nest(dat = c(effect_size, offset0, reject)) %>%
  mutate(mod = map(dat, 
                   ~glm(reject ~ effect_size - 1, 
                        family = binomial(), 
                        data = .x,
                        offset = offset0))) %>%
  mutate(power = map(mod, function(mod) {
    result <- data.frame(effect_size = seq(min_heter_es, max_heter_es, 0.1),
                         offset0 = log(0.05/0.95))
    result$power <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-dat, -mod) %>%
  unnest(power) %>%
  mutate(log_effect_size = log(effect_size)) %>%
  ggplot() +
  geom_line(aes(log_effect_size, power, col = a), size = 1) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  scale_color_manual(values = c(rcartocolor::carto_pal(3, "Earth")[1], "grey", rcartocolor::carto_pal(3, "Earth")[3])) +
  theme_light() +
  theme(legend.position = "bottom") +
  facet_wrap(~conv_or_not) +
  labs(col = "Heteroskedasticity shape")
```

@@

For visual tests, the power curves for normal distribution and lognormal distribution are lower than other two distributions, which are all even. For non-linearity patterns, the power curve of normal distribution is slightly greater than the one for lognormal distribution. And the margin is even larger for heteroskedasticity patterns. This suggests skewness can affect the degree of presence of the visual pattern.

For non-linearity patterns, we do not observe significant difference in power of visual test between uniform distribution and discrete uniform distribution, but the normal distribution and the lognoraml distribution have lower power than the previous two. In terms of conventional tests, the discrete uniform distribution is the one with the greatest power, followed by lognormal, uniform and the normal distribution.

For heteroskedasticity patterns, the power of visual test on lineups with uniform distribution has the greatest power. This is as expected since other three distributions could reduce the chance of revealing the underlying visual pattern because of uneven data points. The power of visual test under the discrete uniform distribution has a relatively great. Considering for heteroskedasticity, the visual patterns are usually detected by connecting the maximum and minimum residuals separately at different fitted values. It will not be greatly affected by the discrete uniform distribution compared to the uniform distribution. Again, visual tests have low power on lineups with the normal distribution and the lognormal distribution. It indicates skewness and unevenness can reduce the detection rate of the visual pattern. In the case of conventional tests, the discrete uniform distribution consistently perform well, followed by uniform, normal and the lognormal distribution.

@@

```{r different-x-dist-poly-power, fig.cap = "Comparison of power on lineups with different fitted value distributions, for conventional tests and visual tests. For visual tests, all the power curves are estimated using five evaluations on each lineup. For lineups with uniform fitted value distribution, the five evaluations are randomly sampled from the total eleven evaluations. The sampling process has been repeated for 500 times and the corresponding power curves are drawn with transparent grey lines. For both types of tests power does change. The discrete fitted value distribution substantially lowers the power of visual tests for non-linearity patterns in residual plots. For heteroskedasticity patterns, normal and lognormal distributions produce lower power for both types of tests.", dev = 'png', dpi = 450}

@@

Now it is important to note that residual plots need to be delivered as a lineup, where it is embedded in a field of null plots. This enables a careful calibration for reading structure in residual plots. XXX More here

@@

As expected, if the fitted value distribution is not uniform, there is a loss of power in the visual test. Structure is hardest to detect if fitted values are lognormal. Also, a surprising find was that the direction of heteroskedasticity appears to affect the ability to visually detect it, with wedge to the right being less detectable.

### 2023-02-15 paper_comparison/paper_comparison/paper_comparison.tex
update the text according to the updated effect size

We make an assumption that all the realizations of \(\boldsymbol{X}\)
are generated from a multivariate normal distribution
\(N(\boldsymbol{0}, 0.3^2\boldsymbol{I})\) and then rescaled by the
min-max normalization. The estimate of \(E\) is obtained by taking the
average of the \(E_{sample}\) calculated for samples drawn from the
given model \(M\) repeatedly.

@@

To get an accurate estimate of the power of conventional tests, we
simulate 200000 samples evenly from the models used in the human subject
experiments, and record the \(p\)-values produced by every sample.

@@

For non-linearity patterns, the power curve of RESET test climbs
aggressively from 7\% to 63\% as \(log_e(E)\) increases from 0 to 2.5,
while power of other tests respond inactively to the change of effect,
showing that RESET test is way more sensitive to the type of model
defects that being considered. Meanwhile, no noticeable residual
depatures can be spotted from the example residual plots.

In terms of heteroskedasticity patterns, the power of Breusch-Pagan test
is also almost always greater than the power of visual test. For
\(log_e(E)\) at around 2.5, where the power curve of the visual test
remains at a low level, the Breusch-Pagan test still have a decent
amount of chance of rejecting \(H_0\). Similarly, the visual feature is
nearly unobservable from the example residual plots.

@@

\caption{The single heteroskedasticity lineup that is rejected by the visual test but not by the BP test. The data plot at panel 17 contains a "Butterfly" shape. It has effect size $ = 4.02$, somewhat surprising that it is not detected by the BP test.}\label{fig:heter-example}
\end{figure}

@@

A primary factor contributes to the non-linearity model is the shape of
the non-linearity pattern. According to Figure
\ref{fig:poly-power-uniform-j}, conventional tests have identical power
in testing the ``U'', the ``S'' and the ``M'' shape. But it has
significant lower power in testing the ``Triple-U'' shape. To understand
why this is, one needs to return to the way the RESET test is applied.
It requires a parameter indicating degree of fitted values to test for,
and the recommendation is to generically use four
(\citet{ramsey_tests_1969}). However, the ``Triple-U'' shape constructed
from the Hermite polynomials use power up to 18. If the RESET test had
been applied using a higher power no less than six, the power curve of
``Triple-U'' shape will have little difference to other shapes. The
recommendation of the polynomial power for the RESET should be revised,
perhaps. This illustrates the sensitivity of conventional testing to the
parameters, and it also points to a limitation that one needs to know
the data generating process in order to set the parameters for the test.

For visual tests, based on the orders of the Hermite polynomials, we
expect the ``U'' shape will be the easiest one to be detected by
subjects followed by the ``S'', ``M'' and ``Triple-U'' shape. From
Figure \ref{fig:poly-power-uniform-j}, it can be observed that the power
curves are mostly aligned with our expectation, except for the ``M''
shape, which is as easy to be detected as the ``U'' shape. This implies
the readability of the shape do not strictly follow the degree of the
polynomials.

@@

(change the butterfly color)

We have also investigated the impact of different heteroskedasticity
shapes on power of conventional tests and visual tests. In theory, the
``Left-triangle'' and the ``Right-triangle'' shapes are functionally
identical from the point of view of a Breusch-Pagan test. As shown in
Figure \ref{fig:heter-power-uniform-a}, this is indeed the case where
little difference between the power curves can be perceived. Similarly,
visual tests should have the smae power in detecting these two shapes if
they are equally likely to be identified. However, it can be observed
from Figure \ref{fig:heter-power-uniform-a} that the power curve for the
``Left-triangle'' shape is constantly higher than the one for the
``Right-triangle'' shape, indicating a potential favour of orientation
by human, which is worth to be explored in future studies. Besides, the
chance of detecting the ``Butterfly'' shape by visual tests are slightly
higher than other two shapes.

>>>>

We make an assumption that all the realizations of \(\boldsymbol{X}\)
are generated from a multivariate normal distribution
\(N(\boldsymbol{0}, 0.3^2\boldsymbol{I})\) and then rescaled by the
min-max normalization. The estimate of \(E\) is obtained by taking the
average of the \(E_{sample}\) calculated for samples drawn from the
given model \(M\) repeatedly. We use the normal distribution as a
general choice to approximate the conditional effect size.

@@

@@

For non-linearity patterns, the power curve of RESET test climbs
aggressively from 7\% to 84\% as \(log_e(E)\) increases from 0 to 2.5,
while power of other tests respond inactively to the change of effect,
showing that RESET test is way more sensitive to the type of model
defects that being considered. Meanwhile, no noticeable residual
departures can be spotted from the example residual plots.

In terms of heteroskedasticity patterns, the power of Breusch-Pagan test
is almost always greater than the power of visual test. For \(log_e(E)\)
at around 2.5, where the power curve of the visual test remains at a low
level, the Breusch-Pagan test has around 50\% chance of rejecting
\(H_0\). Similarly, the visual feature is nearly unobservable from the
example residual plots.

@@

\caption{The single heteroskedasticity lineup that is rejected by the visual test but not by the BP test. The data plot at panel 17 contains a "Butterfly" shape. It has effect size $ = 3.76$, somewhat surprising that it is not detected by the BP test.}\label{fig:heter-example}
\end{figure}

@@

A primary factor contributes to the non-linearity model is the shape of
the non-linearity pattern. According to Figure
\ref{fig:poly-power-uniform-j}, conventional tests have higher power in
testing shapes constructed from lower orders of Hermite polynomials. But
the power in testing the ``Triple-U'' shape drops significantly. To
understand why this is, one needs to return to the way the RESET test is
applied. It requires a parameter indicating degree of fitted values to
test for, and the recommendation is to generically use four
(\citet{ramsey_tests_1969}). However, the ``Triple-U'' shape constructed
from the Hermite polynomials use power up to 18. If the RESET test had
been applied using a higher power no less than six, the power curve of
``Triple-U'' shape will only slight lower than the power curve of ``M''
shape. The recommendation of the polynomial power for the RESET should
be revised, perhaps. This illustrates the sensitivity of conventional
testing to the parameters, and it also points to a limitation that one
needs to know the data generating process in order to set the parameters
for the test.

For visual tests, based on the orders of the Hermite polynomials, we
expect the ``U'' shape will be the easiest one to be detected by
subjects followed by the ``S'', ``M'' and ``Triple-U'' shape. From
Figure \ref{fig:poly-power-uniform-j}, it can be observed that the power
curves are mostly aligned with our expectation, except for the ``M''
shape, which is as easy to be detected as the ``S'' shape. This implies,
unlike the conventional test, the readability of the shape do not
strictly follow the degree of the polynomials.

@@

We have also investigated the impact of different heteroskedasticity
shapes on power of conventional tests and visual tests. In theory, the
``Left-triangle'' and the ``Right-triangle'' shapes are functionally
identical from the point of view of a Breusch-Pagan test. As shown in
Figure \ref{fig:heter-power-uniform-a}, this is indeed the case where
little difference between the power curves can be perceived. Similarly,
visual tests should have the same power in detecting these two shapes if
they are equally likely to be identified. However, it can be observed
from Figure \ref{fig:heter-power-uniform-a} that the power curve for the
``Left-triangle'' shape is higher than the one for the
``Right-triangle'' shape, indicating a potential favour of orientation
by human, which is worth to be explored in future studies. Besides, the
chance of detecting the ``Butterfly'' shape by both conventional and
visual tests are much higher than other two shapes.
